---
title: "Reproducible Research: Assignment 1"
date: "November 10, 2015"
output: 
  html_document: 
    keep_md: yes
---

## 1. Activity Monitoring Data
#### Source: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip

Always write code so that someone will be able to read the code
```{r setoptions, echo = FALSE}
library(knitr)
opts_chunk$set(echo = TRUE)
```



## 2. Loading & Preprocessing Data
### 2.1 Loading Data
```{r}

activity_monitoring_data_URL <-"https://d396qusza40orc.cloudfront.net/repdata/data/activity.zip"
activity_monitoring_data_file <- "activity.zip"
if(!file.exists(activity_monitoring_data_file)) {
  download.file(activity_monitoring_data, destfile = activity_monitoring_data_file)
  unzip(activity_monitoring_data_file)
}

a <- read.csv("activity.csv", sep =",", na.strings = "NA", header = TRUE)

dim(a)
head(a)
```


### 2.2 Process Data

Expected Output: Single Markdown file
process by knitr and transform to HTML
Always include the code generated echo = TRUE

```{r}

CompleteValues <- complete.cases(a)
aa <- a[CompleteValues == T,]

head(aa)  # Cleaned Data
```

## 3. Mean total number of steps taken per day
### 3.1 Total # of steps per day
```{r}
StepsByDay <- aggregate(aa$steps ~ aa$date, FUN = mean)
```

### 3.2 Histogram of the total Number of Steps per day
```{r}
colnames(StepsByDay) <- c("steps.by.day", "steps.mean")
str(StepsByDay)
hist(StepsByDay$steps.mean, 
     main = "Total Number of steps taken each day",
     xlab = "Steps per Day", ylab = "Frequency", col="green")
summary(aa)
```

### 3.3 Report of mean and median of the total number of steps taken per day
```{r}
library(dplyr)
StepsByDay
summarize(StepsByDay, steps.median = median(StepsByDay$steps.mean))

```

## 4. Average Daily Activity Pattern
A time series plot (i.e. type = "l" ) of the 5minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
daily_activity <- aggregate(aa$steps ~ aa$interval, FUN = mean)
colnames(daily_activity) = c("steps.interval", "steps")
par(mar = c(5,4,2,2))
with(daily_activity, plot(steps.interval, steps, type = "l", col = "blue", ylab = "Average Steps Across All Days", xlab = "Interval"))
title(main = "Average Daily Activity Pattern")
sprintf( "5 minute interval, on average across all the days in the dataset, containing the maximum number of steps = %d",
            daily_activity[daily_activity$steps == max(daily_activity$steps),]$steps.interval)
```

## 5. Imputing Missing values
### 5.1 Missing values in the dateset
```{r}
MissingValues <- a[CompleteValues == F,]
sprintf("Total number of missing rows in the data set is %d", nrow(MissingValues))
```

### 5.2 Fill missing values
```{r}
intersect(unique(MissingValues$date),unique(StepsByDay$steps.by.day))
length(intersect(unique(MissingValues$interval),unique(aa$interval)))

StepsByInterval <- with(aa, aggregate(steps ~ interval, FUN = mean))

# Fill Missing values
aaa <- merge(a, StepsByInterval, by = c("interval", "interval"))
colnames(aaa) <- c("interval", "steps", "date", "steps.mean")
new_a <- transmute(aaa, date, interval, steps = ifelse(is.na(steps), steps.mean, steps))
```

### 5.3 Create a new data set with but with the missing data values filled in
```{r}
new_daily_steps <- aggregate(new_a$steps ~ new_a$date, FUN = mean)
new_daily_steps
colnames(new_daily_steps) <- c("date", "steps.mean")
```

### 5.4 Histogram of the total # of steps taken eacy day (missing values filled in)
```{r}
hist(new_daily_steps$steps.mean,
      main = "Total Number of steps taken each day (missing values filled)",
      xlab = "Steps per Day", ylab = "Frequency", col="yellow")
summarize(new_daily_steps, steps.median = median(new_daily_steps$steps.mean))
```

### Impact of imputing missing values
```{r}
par(mfrow = c(1,2), mar = c(5,4,2,2))
hist(StepsByDay$steps.mean, 
     main = "Total Number of steps taken each day",
     xlab = "Steps per Day", ylab = "Frequency", col="green")
hist(new_daily_steps$steps.mean,
      main = "Total Number of steps taken each day (missing values filled)",
      xlab = "Steps per Day", ylab = "Frequency", col="yellow")

```

## 6. Differences in activity patterns between weekdays and weekends
```{r}
day_steps <- transform(new_a, 
          day = ifelse(strftime(date, "%A") %in% c("Sunday", "Saturday"), "weekend", "weekday"))
library(lattice)
par(mar = c(5,4,2,2))
xyplot(steps ~ interval | day, data = day_steps, type = "l",  layout = c(1,2), ylab = "Number of Steps", xlab = "Interval")
```
